---
title: "GSEABenchmarking-Copy"
output: html_document

Adapted from:
title: "Analysis vignette for GSEA Benchmarking manuscript"
author: Ludwig Geistlinger
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  % \VignetteIndexEntry{GSEA Benchmarking manuscript analysis}
  % \VignetteEngine{knitr::rmarkdown}
---

# Setup
```{r packages}
library(EnrichmentBrowser)
library(GSEABenchmarkeR)
library(ggpubr)
```

```{r cb.colors}
cb.pink <- "#CC79A7"
cb.darkred <- "#B42F32"
cb.red <- "#D55E00"
cb.lightred <- "#DF6747"
cb.blue <- "#0072B2"
cb.yellow <- "#F0E442"
cb.green <- "#009E73"
cb.lightblue <- "#56B4E9"
cb.lightorange <- "#FAAC77"
cb.orange <- "#E69F00"
cb.darkorange <- "#F6893D"
cb.lightgrey <- "#C9C9BD"
cb.darkgrey <- "#878D92"
```



# Expression data sources
## Microarray compendium
```{r maComp}
data.dir <- "data"
ma.dir <- file.path(data.dir, "GEO2KEGG_preproc")
geo2kegg <- loadEData(ma.dir)
ma.ids <- names(geo2kegg)
ma.ids
```
## RNA-seq compendium
```{r rseqComp}
rseq.dir <- file.path(data.dir, "TCGA_preproc")
rseq.raw <- file.path(rseq.dir, "GSE62944_matched_limmavoom")
rseq.vst <- file.path(rseq.dir, "GSE62944_matched_vst")
rseq.tpm <- file.path(rseq.dir, "cTD_matched_tpm")
tcga.raw <- loadEData(rseq.raw)
tcga.vst <- loadEData(rseq.vst)
tcga.tpm <- loadEData(rseq.tpm)
rseq.ids <- names(tcga.raw)
rseq.ids
```

# Gene sets
## KEGG
```{r keggGS}
kegg.gs <- getGenesets(org="hsa", db="kegg")
length(kegg.gs)
(min.size <- EnrichmentBrowser::configEBrowser("GS.MIN.SIZE"))
(max.size <- EnrichmentBrowser::configEBrowser("GS.MAX.SIZE"))
ind <- (lengths(kegg.gs) >= min.size) & (lengths(kegg.gs) <= max.size)
kegg.gs <- kegg.gs[ind]
length(kegg.gs)
MASS::truehist(lengths(kegg.gs), nbins=50) 
```
## GO
```{r goGS}
go.gs <- getGenesets(org="hsa", db="go", onto="BP")
length(go.gs)
ind <- (lengths(go.gs) >= min.size) & (lengths(go.gs) <= max.size)
go.gs <- go.gs[ind]
length(go.gs)
MASS::truehist(lengths(go.gs), nbins=50)
```



# Enrichment methods
```{r methods}
ea.methods <- sbeaMethods()[1:10]
```

# Benchmarking
## set dirs
```{r resDir}
res.dir <- file.path(data.dir, "results")
geo.dir <- file.path(res.dir, "GEO2KEGG")

data.dir <- system.file("extdata", package="GSEABenchmarkeR")

kegg.dir <- file.path(geo.dir, "kegg/perm1k")
go.dir <- file.path(geo.dir, "go_bp")

tcga.dir <- file.path(res.dir, "TCGA")
vst.dir <- file.path(tcga.dir, "GSE62944_matched_vst")
rseq.kegg.dir <- file.path(vst.dir, "kegg")
rseq.go.dir <- file.path(vst.dir, "gobp")
```


```{r malaRankings}
mala.kegg.file <- file.path(data.dir, "malacards", "KEGG.rds")
mala.go.file <- file.path(data.dir, "malacards", "GO_BP.rds")

# read RDS files
mala.kegg <- readRDS(mala.kegg.file)
mala.go <- readRDS(mala.go.file)

# show number of rows per disease malacards KEGG
vapply(mala.kegg, nrow, integer(1))

# show number of rows per disease malacards GO
vapply(mala.go, nrow, integer(1))

# optional
#mala.kegg$ALZ
#mala.kegg$BRCA
```

## Mapping between dataset ID and disease code
```{r data2dis}
d2d.file <- file.path(data.dir, "malacards", "GseId2Disease.txt")
d2d.map <- readDataId2diseaseCodeMap(d2d.file)
head(d2d.map)
d2d.tcga <- rseq.ids
names(d2d.tcga) <- rseq.ids
```
```{r readRankings}
ma.kegg.ranks <- readResults(kegg.dir, ma.ids, 
                                methods=ea.methods, type="ranking")
ma.go.ranks <- readResults(go.dir, ma.ids,
                                methods=ea.methods, type="ranking")
lengths(ma.kegg.ranks)

rseq.kegg.ranks <- readResults(rseq.kegg.dir, rseq.ids, 
                                methods=ea.methods, type="ranking")
rseq.go.ranks <- readResults(rseq.go.dir, rseq.ids, 
                                methods=ea.methods, type="ranking")

# optional
#ma.kegg.ranks$ora[1:2]
```

### Here we can calculate the malacard scores! ###

## Relevance score of a gene set ranking
```{r evalRelevance}
ma.kegg.ranks$ora$GSE1297
obs.score <- evalRelevance(ma.kegg.ranks$ora$GSE1297, mala.kegg$ALZ)
obs.score
```
## Theoretical optimum
```{r compOpt}
gs.names <- ma.kegg.ranks$ora$GSE1297$GENE.SET
gs.ids <- substring(gs.names, 1, 8)
opt.score <- compOpt(mala.kegg$ALZ, gs.ids)
opt.score
round(obs.score / opt.score * 100, digits=2)
```

## Random relevance score distribution
```{r compRand}
rand.scores <- compRand(mala.kegg$ALZ, gs.ids, perm=50)
summary(rand.scores)
(sum(rand.scores >= obs.score) + 1) / 51
```


# facetplot function
```{r facetplot}
facetplot <- function(ma.kegg, ma.go, rseq.kegg, rseq.go, 
    ylab="% significant sets", vline=6.5, hline=NA, log=FALSE, orderf=median)
{
    l <- list(ma.kegg=ma.kegg, ma.go=ma.go, rseq.kegg=rseq.kegg, rseq.go=rseq.go)
    df <- reshape2::melt(l)
    gsc <- vapply(df$L1, function(x) unlist(strsplit(x,"\\."))[2], 
                    character(1), USE.NAMES=FALSE)
    df <- cbind(df, gsc=gsc)
    df$gsc <- toupper(df$gsc)
    df$gsc <- vapply(df$gsc, function(n) 
                ifelse(n == "GO", paste(n, "BP", sep="-"), n), 
                character(1), USE.NAMES=FALSE)
    df$gsc <- factor(df$gsc, levels=c("KEGG", "GO-BP"))
    colnames(df)[1:2] <- c("dataset", "method")
    colnames(df)[4] <- "compendium"
    df$compendium <- sub("ma.kegg", "GEO2KEGG microarray", df$compendium)
    df$compendium <- sub("rseq.go", "TCGA RNA-seq", df$compendium)
    df$compendium <- sub("rseq.kegg", "TCGA RNA-seq", df$compendium)
    df$compendium <- sub("ma.go", "GEO2KEGG microarray", df$compendium)
    df$method <- substring(df$method, 1, 7)
    if(log) df$value <- log(df$value, base=10)
    o <- sort(vapply(split(df$value, df$method), 
                        orderf, numeric(1), na.rm=TRUE))
    df$method <- factor(df$method, levels=names(o))
    p <- ggboxplot(df, x = "method", y = "value", 
        width = 0.8, ylab=ylab, xlab="", fill="method")
    p <- ggpar(p, x.text.angle=45, palette = "simpsons", legend="none") 
    if(!is.na(vline)) 
        p <- p + geom_vline(xintercept=vline, linetype="dashed", color = cb.darkgrey)
    if(!is.na(hline)) 
        p <- p + geom_hline(yintercept=hline, linetype="dashed", color = cb.red)

    facet(p, facet.by=c("compendium", "gsc"))    
}    
```

## Cross-dataset relevance score distribution
```{r evalAll, fig.width=6, fig.height=6}
ma.kegg.rel.sets <- evalRelevance(ma.kegg.ranks, mala.kegg, d2d.map)
ma.go.rel.sets <- evalRelevance(ma.go.ranks, mala.go, d2d.map)

rseq.kegg.rel.sets <- evalRelevance(rseq.kegg.ranks, mala.kegg, d2d.tcga)
rseq.go.rel.sets <- evalRelevance(rseq.go.ranks, mala.go, d2d.tcga)
```

# creation of facetplot
```{r}
# facetplot function
facetplot(ma.kegg.rel.sets, ma.go.rel.sets, rseq.kegg.rel.sets, 
    rseq.go.rel.sets, ylab="% optimal relevance score", vline=4.5)
```








